% elliptical-appnote.ltx -- Application note for elliptical functions and integrals on the R47/C47 calculator.

\documentclass[canadian]{article}
\usepackage{amsmath, amssymb, amsthm, graphics, setspace, enumerate, thmtools, tikz, url, parskip, fancyhdr, datetime2}
\usepackage{listings, lastpage}
\usetikzlibrary{angles,quotes}

\fancyhf{}
\fancyhead[L]{\fontsize{8}{10} \selectfont \color{blue}R47/C47 Version 0.109.03.02}
\fancyhead[C]{\fontsize{8}{10} \selectfont \color{blue}Elliptic Functions}
\fancyhead[R]{\fontsize{8}{10} \selectfont \color{blue}\today \ WGR}
\fancyfoot[L]{\fontsize{8}{10} \selectfont \color{blue}Page \thepage \ of \pageref{LastPage}}
\fancyfoot[R]{\fontsize{8}{10} \selectfont \color{blue}R47/C47 Application Note 023}
\pagestyle{fancy}

\lstset{ basicstyle=\ttfamily, mathescape }

\begin{document}

\part*{Application Note}

    This document describes how elliptic integrals and functions are used in the R47/C47 calculator.

    In some cases there is an apparent discrepancy between values the R47/C47 produces, and
    values for the same elliptic functions you may find in tables in books.  This document attempts
    to reconcile those differences.

    Numerical results are given to 5 digits to the right of the decimal point where possible,
    for consistency in comparison.  The text \cite{Lawden01} uses that precision.

\section{Basics of Elliptical Functions and Integrals}

    What are elliptic functions and integrals?

    In some ways they are just solutions to differential equations that have no apparent bearing
    on ellipses at all.

    Here we will explore those issues and show simple examples of computations with elliptic functions
    that you can try on an R47 or C47 calculator, or in Mathematica.

    \subsection{Review:  Basic Geometry of an Ellipse}

        Many people studied conic sections in high school.  See \cite{Rel01} for example.
        A conic section is a geometrical
        idea from ancient Greek mathematics.  Certain curves can be represented as a conic
        section geometrically combined with analytical (Cartesian) geometry.  Examples:

    \begin{table}[ht]
        \caption{Examples of Conic Sections}
        \centering
        \begin{tabular}{c c c}
            \hline\hline
            Name & Equation & Eccentricity \\
            \hline
            Circle & $x^2+y^2=r^2$ & 0 \\
            Parabola & $y^2=4 a x$ & 1 \\
            Ellipse & $\frac{x^2}{a^2}+\frac{y^2}{b^2}=1$ & $\sqrt{1-\frac{b^2}{a^2}}$ \\
            Hyperbola & $\frac{x^2}{a^2}-\frac{y^2}{b^2}=1$ & $\sqrt{1+\frac{b^2}{a^2}}$ \\
        \end{tabular}
    \end{table}

    Since we are interested in ellipses here, let us note that the ellipse has eccentricity
    $e \in [0,1)$.  In the special case where an ellipse has eccentricity 0, it is actually
    a circle.

    \begin{tikzpicture}
        \draw[thick,blue] (0,0) ellipse (2cm and 1cm);
        \draw[->] (-2.5,0) -- (2.5,0) node[right] {$x$};
        \draw[->] (0,-1.5) -- (0,1.5) node[above] {$y$};
        \draw[dashed,red] (0,0) -- (2,0) node[midway,below] {$a$};
        \draw[dashed,red] (0,0) -- (0,1) node[midway,left] {$b$};
        \fill[black] (0,0) circle (2pt) node[below left] {$0$};
        \node[blue, font=\bfseries] at (1,0.5) {Ellipse};
        \coordinate(O) at (0,0);
        \coordinate(A) at (2,0);
        \coordinate(B) at (30:2cm and 1cm);
        \draw (O) -- (B);
        \pic[draw, ->, "$\phi$", angle eccentricity=2.0 ]{angle = A--O--B};
    \end{tikzpicture}

    Besides the equation given above for an ellipse, it can also be represented parametrically:

    $x = a \cos \theta, y = b \sin \theta, \theta \in [0, 2 \pi)$

    \subsection{A Problem to Solve with an Ellipse:  Arc Length and Inverse}

    One sees online comments, e.g. \cite{SE01}, about elliptic integrals and functions having their roots
    in elliptic arc length, but when you dig into textbooks or online articles for
    details, that is not so apparent.  Most of the applications for elliptic integrals
    and functions go back to problems from specialized differential equations or engineering
    problems.

    I persisted in this question and found answers that tie back to the functions
    in the R47/C47 calculator.  Here are my findings.

    First, there is the problem of notation and parameter names.
    This is discussed in \cite{jdcook01}.  An elliptic function like
    $sn(u,m)$ can be described with either $m$ or $k$ given as the second argument, and
    sometimes the second argument is not mentioned at all.

    Similarly, an elliptic integral can be described as $K(m)$ or $F(\phi,k)$ or $F(\phi,m)$,
    with apparently $k$ and $m$ being different things.  And then, their relationship to
    geometric parameters that we know from simple conic descriptions, like $a$ or $b$ or
    $e$, is not obvious.

    The simple answer is, that when you see $k$ as a parameter, it is called the "modulus",
    but actually it is equal to the eccentricity $e$ as seen in elementary conic section
    terminology.  Some sources instead refer to $m$ as the modulus.

    However, though $k$ is often favoured in mathematical treatments, you often
    see in software such as the R47/C47 or Mathematica a parameter called $m$.  Actually, this is
    almost the same thing:  $m=k^2=e^2$.  This $m$ parameter is just the square of
    eccentricity, if you want to track down elliptic arc length calculations.

    Now, a typical second year calculus textbook for engineering students teaches how to
    calculate arc length in two- or three- dimensional Euclidean space, using integration.
    The Olmsted book, \cite{Olmsted01}, does a good job of this.

    To paraphrase \cite{Olmsted01}:
    If $C$ is an arc with parametrization functions $x = f(t), y = g(t)$, with continuous
    derivatives on $[a,b]$, then $C$ is rectifiable with length

    \[L = \int_{a}^{b}{\sqrt{(f'(t))^2+g'(t))^2} d t}\]

    As an example of this, for the ellipse described above, we get (with a little
    rearrangement):

    The total arc length of the ellipse is \cite{Olmsted01}

    \[4 a \int_{0}^{\pi / 2}{\sqrt{1 - k^2 \sin^2{\theta}} d \theta}\]

    This leads to the following:

    Incomplete elliptic integral of the second kind:
    \[E(\phi,m) = \int_{0}^{\phi}{\sqrt{1 - m \sin^2{\theta}} d \theta}\]

    Complete elliptic integral of the second kind:
    \[E(m) = \int_{0}^{2 \pi}{\sqrt{1 - m \sin^2{\theta}} d \theta}\]

    Confusingly, in different parts of the literature, you will see this version:

    Incomplete elliptic integral of the second kind:
    \[E(\phi,k) = \int_{0}^{\phi}{\sqrt{1 - k^2 \sin^2{\theta}} d \theta}\]

    Complete elliptic integral of the second kind:
    \[E(k) = \int_{0}^{2 \pi}{\sqrt{1 - k^2 \sin^2{\theta}} d \theta}\]

    See also \cite{WI01}.

    Just understand that k is eccentricity of the ellipse, $m=k^2$, and that
    for both you have the restriction $k \in [0,1), m \in [0,1)$.

    So now we have an understandable application if you followed all the bits about
    conic sections, eccentricity, and arc length.  We simply want to take an ellipse
    and calculate its arc length.  The incomplete elliptic integral of the
    second kind does this for arbitrary angle $\phi$, meaning we start at angle
    $0$ and rotate counter-clockwise to $\phi$, measuring arc length as we go.
    The complete elliptic integral of the second kind calculates it for the
    whole ellipse, i.e. the perimeter of the ellipse.

    Let us take an example and run it through both Mathematica and the R47/C47.

    Take $a=5, b=1$, so we have the ellipse $\frac{x^2}{5^2}+\frac{y^2}{1}=1$.

    We have eccentricity $k=e=\sqrt{1-\frac{b^2}{a^2}}=\sqrt{24/25}$.  Both Mathematica
    and the R47/C47 use a parameter named $m$, so we use modulus $m=.96$.

    In Mathematica, we can compute this:

    \begin{verbatim}
        EllipticE[.96]

        --> 1.05050
    \end{verbatim}

    As a sanity check, we can also compute:

    \begin{verbatim}
        Integrate[Sqrt[1 - .96 (Sin[t])^2], {t, 0, Pi/2}]

        --> 1.05050
    \end{verbatim}

    Also:

    \begin{verbatim}
        ArcLength[{5 Sin[t], Cos[t]}, {t, 0, Pi/2}]

        --> 5.25251
    \end{verbatim}
    which is $5$ times the value above, also confirming our correctness.

    If we want to calculate the partial arc length, say for an angle $\phi$ of $.2$ radians
    and the same modulus,
    we run

    \begin{verbatim}
        EllipticE[.2, .96]

        --> 0.19872
    \end{verbatim}

    On the R47/C47, we do this:
    \begin{verbatim}
        .96 E(m)

        --> 1.0505
    \end{verbatim}

    For the partial angle:

    \begin{lstlisting}
        .96 Enter .2 $E(\phi,m)$

        --> .19872
    \end{lstlisting}

    This confirms that elliptic integrals of the second kind give correct arc length
    calculations.

    Now let us look at the inverse problem.  For a given modulus and a given arc length,
    we want to know what angle $\phi$ produces that arc length.

    Unfortunately, my researches show that there is no such function in either the mathematical literature or the
    software packages of Mathematica or R47/C47.  We have to resort to other methods such as root
    solving.  See \cite{SE03} for more on this issue.

    In Mathematica, try this:

    \begin{verbatim}
        FindRoot[EllipticE[x, .96] == .19872, {x, .5}]
        --> 0.20000
    \end{verbatim}
    Correct answer.

    In the R47/C47, enter a program:

    \begin{lstlisting}
        LBL EllE
        MVAR 'x'
        MVAR 'm'
        MVAR 'u'
        RCL 'm'
        RCL 'x'
        E($\phi$, m)
        RCL- 'u'
        END
    \end{lstlisting}

    If you run this program \texttt{EllE} under

    \begin{verbatim}
        ADV > SOLVE
    \end{verbatim}

    with a modulus of $.96$ and a $u$ value of $.19872$, you will get an answer
    of $0.20000$.

    This accounts for $E(m)$ and $E(\phi,m)$ on the R47/C47, which are the Elliptic Integrals of the Second Kind.

    \subsection{Elliptic Integrals of the First Kind}

    The integrals of the first kind are $K(m)$ (complete) and $F(\phi, m)$ (incomplete).

    Equations:

    \[F(\phi, m) = \int_{0}^{\phi}{\frac{d \theta}{\sqrt{1-m \sin^2{\theta}}}}\]

    \[K(m) = \int_{0}^{\pi/2}{\frac{d \theta}{\sqrt{1-m \sin^2{\theta}}}}\]

    These also have a geometric interpretation, from Complex Analysis.  On constructing a conformal
    map from an ellipse to a unit circle, the elliptic integral of the first kind
    tells you the angular displacement needed
    to map arc length in the ellipse to the corresponding point in the circle \cite{SE02}.

    In this case, there is an inverse mapping.  $am(x)$, described in
    Section \ref{jacobiEllipFns}, is the inverse
    of the incomplete elliptic integral of the first kind.

    \subsection{Elliptic Integrals of the Third Kind and Jacobi Zeta}

    $\Pi(n,m)$ is the incomplete Elliptic Integral of the third kind.  It has applications in Physics,
    e.g. electrostatic potential.

    In Mathematica, use $EllipticPi[n,m]$.  On the R47/C47, use $\Pi(n,m)$.

    Comparison of values from Mathemetica  and the R47/C47.

    \begin{verbatim}
        EllipticPi[0.4,0.6]

        --> 2.59092
    \end{verbatim}

    \begin{lstlisting}
        .6 ENTER .4 $\Pi(n,m)$

        --> 2.59092
    \end{lstlisting}

    The Jacobi Zeta function is found in Mathematica as $JacobiZeta[\phi,m)$ and on
    the R47/C47 as $Z(\phi,m)$.  It has applications in physics and in interpretation of
    elliptic arc compared with circle.

    Mathematica example:
    \begin{verbatim}
        JacobiZeta[2,.5]

        --> -0.11777
    \end{verbatim}

    R47/C47 Example:
    \begin{lstlisting}
        .5 ENTER 2 $Z(\phi,m)$

        --> -0.11777
    \end{lstlisting}

    These two match.

    \subsection{Jacobi Elliptic Functions} \label{jacobiEllipFns}

    As a foundation to these functions, there are first the four Theta functions.  These arise as solutions
    to a partial differential equation related to heat diffusion over a slab:

    \[\kappa \nabla^{2} \theta = \frac{d \theta}{d t}\]

    These give rise to functions
    $\theta_{1}(z,q), \theta_{2}(z,q), \theta_{3}(z,q), \theta_{4}(z,q)$.

    The R47/C47 calculator does not implement these Theta functions, but Mathematica does:
    $EllipticTheta[a,u,q]$.
    These are used for the theory behind the following Elliptic functions, which also have applications in
    the Elliptic integrals.

    $sn(u,m), cn(u,m), dn(u,m)$.  These are rendered in Mathematica as
    \begin{verbatim}
        JacobiSN[u,m]
        JacobiCN[u,m]
        JacobiDN[u,m]
    \end{verbatim}

    These functions have some interesting identities.  For example:

    $sn(u,m)^2+cn(u,m)^2=1$

    $sn(u,0)=\sin(u)$ for $u$ in radians.

    $cn(u,0)=\cos(u)$ for $u$ in radians.

    We can test these.  For instance, on the R47/C47:

    \begin{verbatim}
        0 ENTER .2 sn(u,m) .2 sin
    \end{verbatim}
    puts matching values ($0.19867$) in the X and Y registers.

    As an example of understanding how to use parameters, the book \cite{Lawden01} shows in
    Table C that for $k=0.5$ and $x=0.8$, $sn(x,k)=0.70421$.

    However, this calculation on the R47/C47:

    \begin{verbatim}
        .5 ENTER .8 sn(u,m)
    \end{verbatim}

    produces the number $.69093$:  an apparent mismatch.

    Is this wrong?  NO.  The calculator got a different answer from the book, but
    both the calculator and the book are right.  WHAT?

    Yes, explaining this issue clearly is one of the main motivations
    for writing this document.  The answer is in the difference in parameter
    interpretation.  At the risk of being too verbose, I \textbf{must} explain
    this point.

    Earlier in this document, I used a lot of space to explain the $k$ parameter
    versus the $m$ parameter.  Some books or software packages expect $k$ as the
    modulus parameter, and some expect $m$ as the modulus parameter.  As explained
    earlier, $m=k^2$.

    Where the book's table has $sn(u,k)$ with $k=0.5$, the calculator wants
    to run $sn(u,m)$ and we have to square the $k$ parameter to get the $m$
    parameter.

    Knowing that $m=k^2$, we
    can recalculate on the R47/C47 this way:

    \begin{lstlisting}
        .5 $x^2$ .8 sn(u,m)
    \end{lstlisting}

    and get the matching value of $0.70421$.

    If this sounds strange, think of trig functions.  The result of the sine function
    depends on whether the
    parameter is interpreted as being in degrees or radians or other units.
    Accordingly, you might have to do a conversion between degrees and radians in
    some cases to make the sine function give the expected result.

    Here, the result
    of $sn()$ depends on whether the second parameter is interpreted as the eccentricity
    or the square of the eccentricity of the ellipse.  Accordingly, you might have to square
    a parameter to make it compatible with the function you are calling.

    More examples comparing $sn()$ calculations with Tables in the back of the \cite{Lawden01}
    textbook:

    \begin{table}[ht]
        \caption{Lawden Comparison Examples: Jacobi Functions}
        \centering
        \begin{tabular}{c c c c c c c c}
            \hline\hline
            k & x & sn x & cn x & dn x & sn(u,m) & cn(u,m) & dn(u,m) \\
            \hline
            .5 & .8 & .70421 & .70999 & .93596 & .70421 & .70999 & .93596 \\
            .6 & 1.3 & .93416 & .35686 & .82816 & .93416 & .35686 & .82816 \\
        \end{tabular}
    \end{table}

    Incomplete Elliptic integrals using degrees:

    \begin{table}[ht]
        \caption{Lawden Comparison Examples: Incomplete Elliptic Integrals}
        \centering
        \begin{tabular}{c c c c c c}
            \hline\hline
            k & $\phi^{\circ}$ & Lawden $F(\phi,k)$ & $F(\phi,k)$ & Lawden $D(\phi,k)$ & $E(\phi,k)$ \\
            \hline
            .4 & 36 & .63459 & .63459 & .62215 & .62215 \\
            .6 & 65 & 1.21431 & 1.21431 & 1.06311 & 1.06311 \\
        \end{tabular}
    \end{table}

    Apart from this issue for $sn()$,
    there is also the Jacobi amplitude function, represented in the R47/C47 as $am(u,m)$
    and in Mathematica as $JacobiAmplitude[u,m]$.
    Here, $u$ is a parameter in radians, and $m$ is the modulus representing
    the square of eccentricity, as before.

    This function is the inverse of the elliptic integral of the first kind.
    To test this on the R47/C47, try an example of $\phi=.1, m=.7$.  Calculation
    of elliptic integral of first kind:

    \begin{lstlisting}
        .7 $x^2$ .1 $F(\phi,m)$

        --> .10008
    \end{lstlisting}

    Calculation of inverse with same modulus:

    \begin{lstlisting}
        .7 $x^2$ .10008 am(u,m)

        --> .10000
    \end{lstlisting}

    \subsection{Summary of R47/C47 Elliptic Functions}

    The following R47/C47 functions have been described:

    \begin{table}[ht]
        \caption{R47/C47 functions described under Ellipt}
        \centering
        \begin{tabular}{c c}
            \hline\hline
            R47/C47 Name & Description \\
            \hline
            $am(u,m)$ & Jacobi Amplitude Function \\
            $sn(u,m)$ & Jacobi sn Function \\
            $cn(u,m)$ & Jacobi cn Function \\
            $dn(u,m)$ & Jacobi dn Function \\
            $K(m)$ & Complete Elliptic Integral of First Kind \\
            $F(\phi,m)$ & Incomplete Elliptic Integral of First Kind \\
            $E(m)$ & Complete Elliptic Integral of Second Kind \\
            $E(\phi,m)$ & Incomplete Elliptic Integral of Second Kind \\
            $\Pi(n,m)$ & Elliptic Integral of Third Kind \\
            $Z(\phi,m)$ & Jacobi Zeta \\
        \end{tabular}
    \end{table}
    In each case except for $\Pi(n,m)$, the $u$ or $\phi$ parameter is an angle in radians, and the
    $m$ parameter is the square of eccentricity of the ellipse.
    For $\Pi(n,m)$, $n$ is the characteristic and $m$ is the square of eccentricity.

    \clearpage

    \begin{thebibliography}{10}
        \bibitem{Lawden01} Lawden, Derek F, \emph{Elliptic Functions and Applications}, Springer-Verlag, 1989
        \bibitem{Olmsted01} Olmsted, John M H, \emph{Advanced Calculus}, Prentice-Hall, 1961
        \bibitem{Rel01} Del Grande and Egsgard, \emph{Relations}, Gage Educational Publishing, 1972
        \bibitem{SE01}
            StackExchange: \emph{What even are elliptic functions?}.
            \url{https://math.stackexchange.com/questions/3383769/what-even-are-elliptic-functions}
            (Accessed:  April 2026).
        \bibitem{jdcook01} Cook, J. D.,
            \emph{Clearing up the confusion around Jacobi functions}
            \url{https://www.johndcook.com/blog/2018/10/12/jacobi-function-nomenclature/}
        \bibitem{SE02}
            StackExchange: \emph{Elliptic integral of the first kind as conformal mapping}.
            \url{https://math.stackexchange.com/questions/2341813/elliptic-integral-of-the-first-kind-as-conformal-mapping}
            (Accessed:  April 2026)
        \bibitem{SE03}
            StackExchange: \emph{Inverse of elliptic integral of second kind}.
            \url{https://math.stackexchange.com/questions/1123360/inverse-of-elliptic-integral-of-second-kind}
            (Accessed:  April 2026)
        \bibitem{WI01}
            Wikipedia:  Elliptic integral
            \url{https://en.wikipedia.org/wiki/Elliptic_integral}
            (Accessed:  April 2026)
    \end{thebibliography}

    \hfill \break

    \subsection*{Author} William Gordon Rutherdale \\ \\
    Copyright (C) 2026 The C47 Development Team. Permission is granted to copy, distribute and/or modify this document
    under the terms of the GNU Free Documentation License, Version 1.3 or any later version published by the Free
    Software Foundation; with no Invariant Sections, no Front-Cover Texts, and no Back-Cover Texts. A copy of the
    license can be downloaded from gnu.org.

\end{document}
